---
title: "Data Analysis"
format: 
  pdf:
    output-file: "data-analysis.pdf"
editor: visual
date: "2026-03-21"
---

## Import

I scraped the 2024-25 NBA Player Stats: Per Game table.

```{r}
#| message: false
library(tidyverse)
library(rvest)

url <- "https://www.basketball-reference.com/leagues/NBA_2025_per_game.html"
html <- read_html(url)
table <- html |> html_element("#per_game_stats") |>
  html_table()
table
```

{{< pagebreak >}}

## Tidy

The rules that make a dataset tidy are:

-   Each variable is a column;each column is a variable

    Holds for my dataset.

-   Each observation is a row; each row is an observation

    Does not hold for my dataset because players who were in more than one team have duplicate rows. I removed the duplicates and retained a summarized row. I also removed the league average row.

-   Each value is a cell; each cell is a single value

    Does not hold for my dataset because the awards column has multiple values per cell. I just removed the column since I don't need it.

```{r}
table <- table |> group_by(Player) |> 
  filter(n() == 1 | str_detect(Team, "TM")) |>
  filter(Player != "League Average") |>
  select(-Awards) |>
  ungroup()

table
```

My dataset is in "wide" format because each row represents one player and each of their metrics is in its own column rather than being spread across multiple rows.

```{r}
# convert "wide" data to "long" format
table |> 
  pivot_longer(
    cols = starts_with("3"), 
    names_to = "3 Point Metric", 
    values_to = "3 Point Metric Values"
  ) |>
  select(
    Player, 
    `3 Point Metric`, 
    `3 Point Metric Values`
  )
```

{{< pagebreak >}}

## Transform

```{r}
# capitalize athlete surnames
library(stringr)

table <- table |> mutate(
  Player = str_replace(Player, " [A-z][^\\s]+", str_to_upper)
  )

table |> select(Player)
```

Date-time manipulation is not applicable to my dataset.

```{r}
# identify any missing values
table |> summarise(across(everything(), function(x) sum(is.na(x))))
```

There are missing values in the FG%, 3P%, 2P%, eFG% and FT% columns. These are N/A because if a player never attempted a FG, 3P, 2P or FT, the calculation would involve dividing by zero.

{{< pagebreak >}}

## Visualize

```{r}
#| message: false
#| warning: false
library(ggrepel)

# convert postion abbreviations to full names
table <- table |>
  mutate(Pos = case_when(
    Pos == "PG" ~ "Point Guard",
    Pos == "SG" ~ "Shooting Guard",
    Pos == "SF" ~ "Small Forward",
    Pos == "PF" ~ "Power Forward",
    Pos == "C" ~ "Center"
  ))

# select a point to annotate in the visualization
annotate <- table |> filter(`eFG%` > 1)

# create visualization
ggplot(
  data = table, 
  mapping = aes(x = PTS, y = `eFG%`, color = MP)) +
  geom_point(alpha = 0.4) +
  geom_point(data = annotate, color = "red") +
  geom_label_repel(
    data = annotate,
    aes(label = str_c("Outlier:\n", Player, "\nFG Attempts: ", FGA)),
    size = 1.5) +
  labs(
    title = "Scoring Volume vs Scoring Efficiency",
    subtitle = "NBA 2024-25 Season",
    caption = "\nData From 
    https://www.basketball-reference.com/leagues/NBA_2025_per_game.html",
    x = "PTS (Scoring Volume)",
    y = "eFG% (Scoring Efficiency)",
    color = "Minutes Played"
  ) +
  theme(plot.caption = element_text(hjust = 0))
```

{{< pagebreak >}}

## Model

Which player metrics have a significant impact on the number of fouls they commit?

```{r}
model <- lm(PF ~ `3PA` + FTA + ORB + 
            STL + BLK + AST + `eFG%` + Pos, 
            data = table) 
summary(model)
```

The three strongest predictors were steals (coef = 0.536, p \< 0.05), offensive rebounds (coef = 0.298, p \< 0.05), and blocks (coef = 0.255, p \< 0.05). Players who steal more, rebound more offensively, and block more shots tend to commit more fouls per game.

{{< pagebreak >}}

## References

Wickham, H., Çetinkaya-Rundel, M., & Grolemund, G. (2023). *R for data science* (2nd ed.). O'Reilly Media. <https://r4ds.hadley.nz/data-tidy.html#widening-data>
